---
title: "Tornados and Floods Are the Storms That Have the Most Impact on Population Health and the Economy Respectively"
output: 
html_document:
keep_md: true
---

<br />

##Synopsis
In this report we aim to show which types of storms that have are most harmful to population health as well as which types of storms have the greatest economic consequences.  To accomplish this we analyzed data downloaded from the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database.  From this data, we found that between 1950 and November 2011, the timeframe that the databased covered, tornadoes were responsible for the greatest cumulative harm on population health, with more deaths and more injuries attributed to tornadoes than any other storm type by a wide margin.  Additionally we found that during the same time period floods were the storm type that had the greatest economic consequences.  Despite the fact that drought was clearly responsible for the greatest cumulative crop damage during the measured period, the property damage associated with flooding caused floods to dwarf drought when analyzing the total economic impact of both property damage and crop damage combined.

<br />

## Data Processing
Loading Data:  
We load the data into R that was downloaded from the NOAA storm database, decompressing the file if it has not been done so already.
```{r cache=TRUE, message=FALSE}
##check to see if data is decompressed, if it isn't then decompress it
library(R.utils)
if (!file.exists("repdata-data-StormData.csv")){   
        bunzip2("repdata-data-StormData.csv.bz2")
}
data <- read.csv("repdata-data-StormData.csv")
head(data)
```

<br />
Data Processing:  
1. Select only the EVTYPE column and the columns relating to physical health and economic damage. (EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP).  
2. Change all names, as well as the values of the exponent fields, to lowercase to make them easier to work with.  
3. Replace the exponent codes with the proper multiplier and then multiply those multipliers by the observed values.  
4. Select only the evtype, fatalities, injuries, propDamage, and cropDamage columns.
```{r cach=TRUE, message=FALSE, warning=FALSE}
##select columns using select function in dplyr library
library(dplyr)
data <- select(data, EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP)

## change names to lowercase
names(data) <- tolower(names(data))

## change exponent fields to lowercase
data$propdmgexp <- tolower(data$propdmgexp)
data$cropdmgexp <- tolower(data$cropdmgexp)

## replace the exponents with numeric values
data$propdmgexp <- gsub("h", "100", data$propdmgexp)
data$propdmgexp <- gsub("k", "1000", data$propdmgexp)
data$propdmgexp <- gsub("m", "1000000", data$propdmgexp)
data$propdmgexp <- gsub("b", "1000000000", data$propdmgexp)

data$cropdmgexp <- gsub("h", "100", data$cropdmgexp)
data$cropdmgexp <- gsub("k", "1000", data$cropdmgexp)
data$cropdmgexp <- gsub("m", "1000000", data$cropdmgexp)
data$cropdmgexp <- gsub("b", "1000000000", data$cropdmgexp)

data$propdmgexp <- as.numeric(data$propdmgexp)
data$cropdmgexp <- as.numeric(data$cropdmgexp)

##add propdamage and cropdamage fields
data <- mutate(data, propDamage = propdmg * propdmgexp, cropDamage = cropdmg * cropdmgexp)

##subset the dataset to only exclude the propdmg, propdmgexp, cropdmg, 
##cropdmgexp columns
data <- select(data, evtype, fatalities, injuries, propDamage, cropDamage)

head(data)
```

<br />
5. Group data based on event type.
```{r}
##group data by evtype, find number of injuries and fatalities for each event
groupedData <- group_by(data, evtype)

head(groupedData)
```

<br />
6. Process and subset data for effects on public health: get cumulative sum for each column for each event, add column showing sum of both injuries and fatalities, convert evtype values to title case.
```{r}
healthEffect <- summarise(groupedData, deaths=sum(fatalities, na.rm=TRUE), injured=sum(injuries, na.rm=TRUE))

##add column 'incidents' that is total of both injuries and fatalities
healthEffect <- mutate(healthEffect, incidents = deaths + injured)

##convert event types to title case
events <- healthEffect[,1]
events <- gsub("\\b([A-Z])([A-Z]+)", "\\U\\1\\L\\2" ,events, perl=TRUE)
healthEffect[,1] <- events

head(healthEffect)
```

<br />
7. Process and subset data for economic impact: get cumulative sum for each column for each event, add column showing sum of both property damage and crop damage, convert evtype values to title case.
```{r}
ecoEffect <- summarise(groupedData, propDamage=sum(propDamage, na.rm=TRUE), cropDamage=sum(cropDamage, na.rm=TRUE))

##add column 'incidents' that is total of both injuries and fatalities
ecoEffect <- mutate(ecoEffect, totalDamage = propDamage + cropDamage)

##convert event types to title case
events <- ecoEffect[,1]
events <- gsub("\\b([A-Z])([A-Z]+)", "\\U\\1\\L\\2" ,events, perl=TRUE)
ecoEffect[,1] <- events

head(ecoEffect)
```

<br />
<br />

##Results
###Population Health
Find the top 10 events for incidents, defined by the sum of injuries and fatalities.
```{r message=FALSE}
##subset to the top 10 incidents, melt table so it can be plotted
library(reshape)
n <- 10
topIncidents <- arrange(healthEffect, desc(incidents))[1:n,]
plotData <- as.data.frame(melt(topIncidents))

##replace variables for plot
plotData$variable <- gsub("deaths", "Fatalities", plotData$variable)
plotData$variable <- gsub("incidents", "Fatalities & Injuries", plotData$variable)
plotData$variable <- gsub("injured", "Injuries", plotData$variable)

head(plotData)
```

```{r message=FALSE, fig.width = 8, fig.height= 8}
##load gridExtra library, need to plot as 3 charts rather than normal facets so y
##axis is more readable
library(gridExtra)
library(ggplot2)
library(scales)

g1 <- ggplot(subset(plotData, variable == "Fatalities & Injuries"), aes(evtype, value))
g1 <- g1 + geom_bar(stat="identity") + facet_wrap(~variable)
g1 <- g1 + ggtitle("Effects on Population Health by Event Type") 
g1 <- g1 + xlab("Event") + ylab("Count")
g1 <- g1 + theme(text = element_text(size=9))

g2 <- ggplot(subset(plotData, variable == "Fatalities"), aes(evtype, value))
g2 <- g2 + geom_bar(stat="identity") + facet_wrap(~variable)
g2 <- g2 + xlab("Event") + ylab("Count")
g2 <- g2 + theme(text = element_text(size=9))

g3 <- ggplot(subset(plotData, variable == "Injuries"), aes(evtype, value))
g3 <- g3 + geom_bar(stat="identity") + facet_wrap(~variable)
g3 <- g3 + xlab("Event") + ylab("Count")
g3 <- g3 + theme(text = element_text(size=9))

grid.arrange(g1,g2,g3)
```
<br />
As we can see from the above figure, Tornadoes are responsible for the most fatalities as well as the most injuries with excessive heat coming in at a distant second for both categories.

<br />
<br />

###Economic Consequences   
Find the top 10 events for total damage, defined as the sum of crop and property damage.
```{r message=FALSE}
##subset to the top 10 incidents, melt table so it can be plotted
n <- 10
topEco <- arrange(ecoEffect, desc(totalDamage))[1:n,]
plotData <- as.data.frame(melt(topEco))

##replace variables for plot
plotData$variable <- gsub("cropDamage", "Crop Damage $", plotData$variable)
plotData$variable <- gsub("propDamage", "Property Damage $", plotData$variable)
plotData$variable <- gsub("totalDamage", "Total Damage $", plotData$variable)

head(plotData)
```

```{r message=FALSE, fig.width = 8, fig.height= 8}
##need to plot as 3 charts rather than normal facets so y axis is more readable

g1 <- ggplot(subset(plotData, variable == "Total Damage $"), aes(evtype, value))
g1 <- g1 + geom_bar(stat="identity") + facet_wrap(~variable)
g1 <- g1 + ggtitle("Economic Impact by Event Type") 
g1 <- g1 + xlab("Event") + ylab("Dollars")
g1 <- g1 + scale_y_continuous(labels=dollar)
g1 <- g1 + theme(text = element_text(size=9))

g2 <- ggplot(subset(plotData, variable == "Property Damage $"), aes(evtype, value))
g2 <- g2 + geom_bar(stat="identity") + facet_wrap(~variable)
g2 <- g2 + xlab("Event") + ylab("Dollars")
g2 <- g2 + scale_y_continuous(labels=dollar)
g2 <- g2 + theme(text = element_text(size=9))

g3 <- ggplot(subset(plotData, variable == "Crop Damage $"), aes(evtype, value))
g3 <- g3 + geom_bar(stat="identity") + facet_wrap(~variable)
g3 <- g3 + xlab("Event") + ylab("Dollars")
g3 <- g3 + scale_y_continuous(labels=dollar)
g3 <- g3 + theme(text = element_text(size=9))

grid.arrange(g1,g2,g3)
```
<br />
As we can see from the above figure, floods were resposible for the most cumulative total damage of both property and crops over the measured period.  However, when looking exclusively at crops, drought was the event that had the greatest economic impact.

