---
title: "Analyzing the Storms:  Their Effect on Population Health and Economic Consequences"
output: 
html_document:
keep_md: true
---

<br />
<br />

##Synopsis
Text

<br />
<br />

## Data Processing
We load the data into R that was downloaded from the NOAA storm database, decompressing the file if necessary.
```{r cache=TRUE, message=FALSE}
##check to see if data is decompressed, if it isn't then decompress it
library(R.utils)
if (!file.exists("repdata-data-StormData.csv")){   
        bunzip2("repdata-data-StormData.csv.bz2")
}
data <- read.csv("repdata-data-StormData.csv")
head(data)
```

Select only the EVTYPE column and the columns relating to physical health and economic damage. (EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP).  Change all names to lowercase to make them easier to work with.

```{r cach=TRUE, message=FALSE, warning=FALSE}
##select columns using select function in dplyr library
library(dplyr)
data <- select(data, EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP)

## change names to lowercase
names(data) <- tolower(names(data))

## change exponent fields to lowercase
data$propdmgexp <- tolower(data$propdmgexp)
data$cropdmgexp <- tolower(data$cropdmgexp)

## replace the exponents with numeric values
data$propdmgexp <- gsub("h", "100", data$propdmgexp)
data$propdmgexp <- gsub("k", "1000", data$propdmgexp)
data$propdmgexp <- gsub("m", "1000000", data$propdmgexp)
data$propdmgexp <- gsub("b", "1000000000", data$propdmgexp)

data$cropdmgexp <- gsub("h", "100", data$cropdmgexp)
data$cropdmgexp <- gsub("k", "1000", data$cropdmgexp)
data$cropdmgexp <- gsub("m", "1000000", data$cropdmgexp)
data$cropdmgexp <- gsub("b", "1000000000", data$cropdmgexp)

data$propdmgexp <- as.numeric(data$propdmgexp)
data$cropdmgexp <- as.numeric(data$cropdmgexp)

##add propdamage and cropdamage fields
data <- mutate(data, propDamage = propdmg * propdmgexp, cropDamage = cropdmg * cropdmgexp)

##subset the dataset to only exclude the propdmg, propdmgexp, cropdmg, 
##cropdmgexp columns
data <- select(data, evtype, fatalities, injuries, propDamage, cropDamage)

head(data)
```

Group data based on event type.
```{r}
##group data by evtype, find number of injuries and fatalities for each event
groupedData <- group_by(data, evtype)

groupedData
```

Process and subset data for effects on public health.
```{r}
healthEffect <- summarise(groupedData, deaths=sum(fatalities, na.rm=TRUE), injured=sum(injuries, na.rm=TRUE))

##add column 'incidents' that is total of both injuries and fatalities
healthEffect <- mutate(healthEffect, incidents = deaths + injured)

##convert event types to title case
events <- healthEffect[,1]
events <- gsub("\\b([A-Z])([A-Z]+)", "\\U\\1\\L\\2" ,events, perl=TRUE)
healthEffect[,1] <- events

healthEffect
```

Process and subset data for economic impact.
```{r}
ecoEffect <- summarise(groupedData, propDamage=sum(propDamage, na.rm=TRUE), cropDamage=sum(cropDamage, na.rm=TRUE))

##add column 'incidents' that is total of both injuries and fatalities
ecoEffect <- mutate(ecoEffect, totalDamage = propDamage + cropDamage)

##convert event types to title case
events <- ecoEffect[,1]
events <- gsub("\\b([A-Z])([A-Z]+)", "\\U\\1\\L\\2" ,events, perl=TRUE)
ecoEffect[,1] <- events

ecoEffect
```

##Results
Find the top 10 events for incidents, defined by the sum of injuries and fatalities.
```{r message=FALSE}
##subset to the top 10 incidents, melt table so it can be plotted
library(reshape)
n <- 10
topIncidents <- arrange(healthEffect, desc(incidents))[1:n,]
plotData <- as.data.frame(melt(topIncidents))

##replace variables for plot
plotData$variable <- gsub("deaths", "Fatalities", plotData$variable)
plotData$variable <- gsub("incidents", "Fatalities & Injuries", plotData$variable)
plotData$variable <- gsub("injured", "Injuries", plotData$variable)

head(plotData)
```

```{r message=FALSE, fig.width = 8, fig.height= 8}
## load gridExtra library, need to plot as 3 charts rather than normal facets so y
## axis is more readable
library(gridExtra)
library(ggplot2)

g1 <- ggplot(subset(plotData, variable == "Fatalities & Injuries"), aes(evtype, value))
g1 <- g1 + geom_bar(stat="identity") + facet_wrap(~variable)
g1 <- g1 + ggtitle("Effects on Population Health by Event Type") 
g1 <- g1 + xlab("Event") + ylab("Count")
g1 <- g1 + theme(text = element_text(size=9))

g2 <- ggplot(subset(plotData, variable == "Fatalities"), aes(evtype, value))
g2 <- g2 + geom_bar(stat="identity") + facet_wrap(~variable)
g2 <- g2 + xlab("Event") + ylab("Count")
g2 <- g2 + theme(text = element_text(size=9))

g3 <- ggplot(subset(plotData, variable == "Injuries"), aes(evtype, value))
g3 <- g3 + geom_bar(stat="identity") + facet_wrap(~variable)
g3 <- g3 + xlab("Event") + ylab("Count")
g3 <- g3 + theme(text = element_text(size=9))

grid.arrange(g1,g2,g3)
```

Find the top 10 events for total damage, defined as the sum of crop and property damage.
```{r message=FALSE}
##subset to the top 10 incidents, melt table so it can be plotted
n <- 10
topEco <- arrange(ecoEffect, desc(totalDamage))[1:n,]
plotData <- as.data.frame(melt(topEco))

##replace variables for plot
plotData$variable <- gsub("cropDamage", "Crop Damage $", plotData$variable)
plotData$variable <- gsub("propDamage", "Property Damage $", plotData$variable)
plotData$variable <- gsub("totalDamage", "Total Damage $", plotData$variable)

head(plotData)
```

```{r message=FALSE, fig.width = 8, fig.height= 8}
## load gridExtra library, need to plot as 3 charts rather than normal facets so y
## axis is more readable

g1 <- ggplot(subset(plotData, variable == "Total Damage $"), aes(evtype, value))
g1 <- g1 + geom_bar(stat="identity") + facet_wrap(~variable)
g1 <- g1 + ggtitle("Economic Impact by Event Type") 
g1 <- g1 + xlab("Event") + ylab("Dollars")
g1 <- g1 + theme(text = element_text(size=9))

g2 <- ggplot(subset(plotData, variable == "Property Damage $"), aes(evtype, value))
g2 <- g2 + geom_bar(stat="identity") + facet_wrap(~variable)
g2 <- g2 + xlab("Event") + ylab("Dollars")
g2 <- g2 + theme(text = element_text(size=9))

g3 <- ggplot(subset(plotData, variable == "Crop Damage $"), aes(evtype, value))
g3 <- g3 + geom_bar(stat="identity") + facet_wrap(~variable)
g3 <- g3 + xlab("Event") + ylab("Dollars")
g3 <- g3 + theme(text = element_text(size=9))

grid.arrange(g1,g2,g3)
```
<br />
<br />


